Handling missing data in smartphone location logs

### 2017-11-27 thesisReport.Rmd
new plots, scripts and added to the text

How people move about in their environment affects a wide range of outcomes. These include health, income and social capital [@goodchild_toward_2010]. A better understanding of mobility could lead to better health and urban-planning policies. A large part of studies on human mobility are conducted with pen-and-paper travel diaries. These surveys have known methodological flaws. The high cost and burden to respondents limits the period of data collection. Short trips are often under reported  [@wolf_impact_2003] . The self-reported duration of commutes is often underestimated [@delclos-alio_keeping_2017].

@@

In most studies participants receive a specialised GPS devices to track their mobility.  We call resulting logs *specialised logs*. @barnett_inferring_2016 point out several methodological issues with these studies. High costs and a high burden to participants limits their scalability. Besides, introducing a new device to the participant's life may bias the data. Due to these drawbacks specialised logs usually span a short amount of time. @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*). Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) . For instance,  Google Location History contains information on millions of users [@location_history_timeline_2017]. Often, secondary logs span several years. By law, secondary logs are accessible to users for free [@commission_protecting_2017]. Yet, secondary logs also present methodological challenges. They were created for non-academic purposes under engineering constraints. These constraints mean that sensors do not track users continuously. The resulting logs can be sparse and inaccurate. Hence, two important challenges are dealing with measurement noise and missing data.

Missing data is a pervasive issue as it can arise due to several reasons. Technical reasons include signal loss, battery failure and device failure. Behavioural reasons include leaving the phone at home or switching the device off. As a result, secondary logs often contain wide temporal gaps with no measurements. For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017]. Other researchers report similar trends in other fields [e.g. @harari_using_2016;@jankowska_framework_2015].

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Traditional methods, such as mean imputation, are not suitable for spatiotemporal data. This is because measurements are often correlated in time and space. For example, imagine an individual who splits almost all her time between work or at home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work.  She has never and will never be there! Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

Smartphone location measurements are substantially less accurate than those of professional GPS trackers. Android phones collect location information through a variety of less accurate methods. Other than GPS measurements, Androids use WiFi access points and cellphone triangulation. Different methods are used because of computational and battery constraints [@lamarca_place_2005; @chen_practical_2006]. In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time.

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. If the measurements are frequent and noisy, we will calculate a lot of movement. Even if the individual did not move at all! The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

In this paper we will compare methods used to deal with measurement error and missing data in mobility patterns from secondary GPS logs.

@@

Given that there is little literature on dealing with missing data in custom or secondary logs it is worth illustrating the typical characteristics of this data using an example data set. The example dataset comes from the Google Location History of a single individual and spans from January 2013 to January 2017. It was recorded with multiple different Android devices and contains 814 941 measurements, with approximately 742 measurements per day ($\widehat{\sigma}$=868.15). The dataset contains a wide range of variables including inferred activity and velocity. For the purposes of this paper we will focus only on latitude, longitude, accuracy (defined below) and a timestamp.

## Location logs and notation

GPS measurements report our location on a three dimensional planet, yet we are interested in placing these measurements on a two dimensional map. Projecting three dimensional measurements onto a two dimensional plane results in errors, in order to minimise these errors we borrow an error minimising projection method from @barnett_inferring_2016.

Let a persons' true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively. Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude. In addition, let $a$ denote the estimated accuracy of the recorded data.  accuracy. $G(t)$, $D$ and $a$ are indexed by time labled by the set $T = t_1 < ... < t_{n+1}$. For simplicity, let each entry in the discrete index set $T$ represent a 5 minute window. The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle. If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.

@@

In the example data set the distribution of $a$ is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km. @palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. We observe similar patterns in secondary logs. Figure 1 shows how accuracy can vary as a function of user behaviour, time and location. Inaccurate measures are often followed by more accurate measures.  Most notably, low accuracy often (but not always) is  associated with movement (Figure 2). Stationary accuracy varies depending on phone battery level, wifi connection and user phone use. There are several recurring low-accuracy points, possibly the result of cell-phone tower triangulation.

```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement on a morning journey. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. ",out.width = '100%'}
include_graphics("img/journeyTillMiddayBoaz.png")
```

```{r, accuracyPlot2, echo=FALSE, fig.cap="The upper chart shows the rolling mean distance from the next point in meters over the course of the day. The lower chart shows the rolling mean accuracy over the course of the day. In both cases, the rolling window is 3.",out.width = '100%'}
include_graphics("img/accuracyLocShift.png")

@@

Over 54% of the data is missing for the entire duration of the log (as shown in Figure X a). However,this is misleading as there are several long periods with no measurements whatsoever. The structure of missingness of a day with measurements is shown in Figure X b. As you can see, there are several long periods over the course of the log for which there are no measurements. Moreover, even during a single day there are continuous periods where there is missing data, mostly during the late hours of the night in this case.

@@

```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data over a day. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.",out.width = '100%'}
#fix width
include_graphics("img/missingBoaz5minExample.png")
```

## Current methods for imputing missing spatiotemporal data

As we have mentioned before, @jankowska_framework_2015 point out that there is little transparency regarding decisions of how to deal with missing data in GPS mobility logs. We suspect this is because the highly interdisciplinary nature of the problem means researchers are unaware of potential solutions. For this reason, we consider it important to briefly discuss a few methods one could consider in order to deal with measurement inaccuracy and missing data problems in spatiotemporal data. In doing so, we will argue that state space models (SSMs) and spatiotemporal imputation methods, which are extensively used to model missing and noisy spatiotemporal data, are not well suited to deal with human mobility patterns. Moreover, we discuss in detail two approaches by @palmius_detecting_2017 and @barnett_inferring_2016 which deal explicitly with missing data in mobility patterns from smartphone GPS logs.

There is a vast literature of using SSMs to improve measurements accuracy and deal with missing data. Behavioural ecologists for instance, have used SSMs extensively to explain how animals interact with their environment [@patterson_statespace_2008]. These models can be quite complex, for example @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements. The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems. In fact, the extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013]. The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.

However, the main limitation of SSMs is that they ignore the fact that humans have regular movement routines, such as going to work or shopping for groceries on weekends. This limitation is due to the fact that SSMs are based on the Markov property. Thus, the estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$. Hierarchical structuring and conditioning on a larger context have been suggested as ways to improve the performance of Markovian models, but these solutions are often computationally intractable or unfeasible [@sadilek_far_2016]. For this reason we do not consider SSMs to be useful for imputing missing data, although they could be of use in filtering noise.

In addition to SSMs there are spatiotemporal imputation methods often used in climate or geological research for estimating missing data. For instance, @feng_cutoff:_2014 illustrate their CUTOFF method, which relies on estimating missing values using the nearest observed neighbours in time, using rainfall data from dozens of gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to present their model based on underground water data in China.

The difficulty of applying this class of spatiotemporal imputation models to human mobility tracks lies in the fact that these models generally assume fixed measurement stations (such as rainfall gauging stations). While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostensibly by dividing the sample space into rasters analogous to measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible. To our knowledge such models have not been implemented for mobility traces.

@@

```{r datadetailsTable, echo = F}
datadetails <- readRDS("tempdata/datadescriptives.rds")
datadetails <- datadetails[,-7] #remove last column for now
kable(datadetails,digits = 2,col.names = c("Log duration",
                                           "Logged days",
                                           "Observations",
                                           "Missing days",
                                           "Missing data",
                                           "Mean Accuracy"),
      caption = "Table with descriptives about the data sets used to build the imputation methods. ")
```

@@

Analyses were performed using R and a multitude of other statistical packages [@ggplot2;@dplyr;@leaflet;@base;@sp1;@sp2].

@@

## References

>>>>

How people move about in their environment affects a wide range of outcomes, such as  health, income and social capital [@goodchild_toward_2010]. A better understanding of mobility could lead to better health and urban-planning policies. Yet a large part of studies on human mobility are conducted with pen-and-paper travel diaries, despite well known methodological flaws. The high cost and burden to respondents limits the span of data collection. Short trips are frequently under-reported  [@wolf_impact_2003] . The self-reported duration of commutes is often underestimated [@delclos-alio_keeping_2017].

@@

In most studies participants receive a specialised GPS devices to track their movement. We call resulting logs *specialised logs*. @barnett_inferring_2016 point out several methodological issues with these studies. Like studies using pen-and-paper travel diaries, collecting specialised logs is costly and places a high burden on participants. Besides, introducing a new device to the participant's life may bias their behaviour. Due to these drawbacks specialised logs usually span a short amount of time. @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*). Another solution is to take advantage of existing smartphone location logs  (*secondary logs*) . For instance,  Google Location History contains information on millions of users [@location_history_timeline_2017]. Often, secondary logs span several years. By law, secondary logs are accessible to users for free [@commission_protecting_2017]. Yet, secondary logs also present methodological challenges. They were created for non-academic purposes under engineering constraints (detailed subsequently). These constraints mean that sensors do not track users continuously, meaning that the resulting logs can be sparse and inaccurate. Hence, two important challenges are dealing with measurement noise and missing data.

Missing data is a pervasive issue as it can arise due to several reasons. Technical reasons include signal loss, battery failure and device failure. Behavioural reasons include leaving the phone at home or switching the device off. As a result, secondary logs often contain wide temporal gaps with no measurements. For instance, several research groups studying mental health report missing data rates between 30% to 50% [@saeb_mobile_2015;@grunerbl_smartphone-based_2015;@palmius_detecting_2017]. Other researchers report similar trends in different fields [e.g. @harari_using_2016;@jankowska_framework_2015].

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Importantly, spatiotemporal data measurements are often correlated in time and space. This means that common methods, such as mean imputation, are unsuitable.  For example, imagine an individual who splits almost all her time between work and home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work.  She has never and will never be there! Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

The accuracy of smartphone location measurements is substantially lower than that of professional GPS trackers. Android phones collect location information through a variety of methods. Other than GPS measurements, Androids use less-accurate heuristics such as WiFi access points and cellphone triangulation. Different methods are used because of computational and battery constraints. GPS is the most energy consuming sensor on most smartphones [@lamarca_place_2005; @chen_practical_2006]. In professional GPS trackers less than 80% of measurements fall within 10 meters of the true location. GPS measures are most inaccurate in dense urban locations and indoors [@schipperijn_dynamic_2014;@duncan_portable_2013]. Unfortunately for researchers, this is where people in the developed world spend most of their time.

Noisy data can lead to inaccurate conclusions if it is not accounted for. Suppose we wish to calculate an individual's movement in a day. A simple approach would be to calculate the sum of the distance between each measurement. But if there is noise, the coordinates will vary even though the individual is not moving. If the measurements are frequent and noisy, we will calculate a lot of movement, even if the individual did not move at all! The problem is further complicated because missing data and noisy measurements are related. Methods used by researchers to reduce noise, such as throwing out inaccurate measurements [e.g. @palmius_detecting_2017],  can exacerbate the severity of the missing data problem.

In this paper we will explore in detail the problem of missing data and measurement error in secondary location logs. Moreover, we will  compare methods used to deal with these problems.

@@

There is little literature on dealing with missing data in custom or secondary logs. Thus it is worth illustrating the typical characteristics using an example data set. The example data set comes from the Google Location History of a single individual. It spans from January 2013 to January 2017 and contains  814 941  measurements.  The data set contains a multitude of variables, including inferred activity and velocity. We will focus on measurements of latitude, longitude, accuracy (defined below). All measurements are paired with a timestamp.

@@

Google location history provides a measure of accuracy that is given in meters such that it represents the radius of a 67% confidence circle. In the example data set the distribution of accuracy is highly right skewed, with a median of 28, $\mu = 127$ and the maximum value at 26 km. @palmius_detecting_2017 note that in their Android based custom logs inaccurate location values are interspersed between more accurate location values at higher sample rates per hour. We observe similar patterns in secondary logs. Figure 1 shows how accuracy can vary as a function of user behaviour, time and location. Inaccurate measures are often followed by more accurate measures.  Most notably, low accuracy often (but not always) is  associated with movement (Figure 2). Stationary accuracy varies depending on phone battery level, wifi connection and user phone use. There are several recurring low-accuracy points, possibly the result of cell-phone tower triangulation.

```{r, accuracyPlot,echo = FALSE,fig.cap="Measurement accuracy of each logged measurement of a morning journey on February 15th 2017. This includes all measurements from midnight to midday. The red circles denote the accuracy of all logged measurement points (the raw data). The points connected in time are connected by a line. The blue line shows the path without the most inaccurate (accuracy > 400 meters) points filtered out. The red line shows the path with all measurements included. ",out.width = '100%'}
include_graphics("img/journeyTillMiddayBoaz.png")
```

```{r, accuracyPlot2, echo=FALSE, fig.cap="Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the rolling mean distance from the next point in meters over the course of the day. The first peak corresponds to the first journey from the users home a gym around 8am. The second, smaller peak before 10 reflects a journey from the gym to the lecture theatre. Both journies can be seen in Figure 1. The lower chart shows the rolling mean accuracy over the course of the day. In both cases, the rolling window is 3. The figure shows that measurement inaccuracy is sometimes related to the movement of the individual.",out.width = '100%'}
include_graphics("img/accuracyLocShift.png")

@@

Over 54% of the data is missing for the entire duration of the log. This may be misleading as there are several long periods with no measurements whatsoever (see Figure X a). For days which were not entirely missing, approximately 22% of all five minute segments were missing. The structure of missingness of a day with measurements is shown in Figure X b. As you can see, there are several long periods over the course of the log for which there are no measurements. Moreover, even during a single day there are continuous periods where there is missing data, mostly during the late hours of the night in this case.

@@

```{r, measurementsPerDay,echo = FALSE,fig.cap="Example of missing data on February 15th 2017. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.",out.width = '100%'}
#fix width
include_graphics("img/missingBoaz5minExample.png")
```

## Two-dimensional projections and notation

GPS measurements provide us with coordinates on the surface of the earth. Because most mobility metrics are computed for data in $\mathbb{R}^2$,  we are interested in mapping these $\mathbb{R}^3$ measurements on a 2D Euclidean plane. Projecting three dimensional measurements onto a two dimensional plane results in distortion. To minimise errors we borrow an error minimising projection method from @barnett_inferring_2016.

Having thus converted lattitude and longitude onto coordinates unique to each individual, let a person's true location on this two-dimensional plane be $G(t) = [G_x(t) G_y(t)]$ where $G_x(t)$ and $G_y(t)$ denote the location of the individual at time $t$ on the x-axis and y-axis respectively. Moreover, let $D \in \mathbb{R}^2$ be the recorded data containing lattitude and longitude. In addition, let $a$ denote the estimated accuracy of the recorded data.  accuracy. $G(t)$, $D$ and $a$ are indexed by time labled by the set $T = t_1 < ... < t_{n+1}$. For simplicity, let each entry in the discrete index set $T$ represent a 5 minute window. The measure of accuracy $a_t$ is given in meters such that it represents the radius of a 67% confidence circle. If $D_t = \emptyset$ it is considered *missing* and it is not missing otherwise.

## Selecting candidate models

There is no golden standard or established practice in how to deal with missing data in GPS logs. Researchers are generally vague about what practices they follow [@jankowska_framework_2015]. Ostensibly this is because they are unaware of possible solutions. In an attempt to elucidate the topic, we explore potential solutions.  We will argue that extensively used spatiotemporal methods, such as state space models (SSMs), are not well suited to deal with human mobility patterns. We also discuss in detail two approaches which deal explicitly with mobility patterns from custom or secondary logs [ @palmius_detecting_2017 ;@barnett_inferring_2016].

There is a vast literature on using SSMs in spatiotemporal statistics. For example, ecologists have used SSMs to explain how animals interact with their environment [@patterson_statespace_2008]. These models can be quite complex. @preisler_modeling_2004 uses Markovian movement processes to characterise the effect of roads, food patches and streams on cyclical elk movements. The most well studied SSM is the Kalman filter, which is the optimal algorithm for inferring linear Gaussian systems. The extended Kalman filter is the de facto standard for GPS navigation [@chen_state_2013]. The advantage of state space models is that they are flexible, deal with measurement inaccuracy, include information from different sources and can be used in real time.

For us, the main limitation of SSMs is that they ignore regular movement routines.  For instance, humans tend to go to work on weekdays and sleep at night. Because SSMs are based on the Markov property, they cannot incorporate this information. The estimated location $G(t)$ at timepoint $t$ is often based only upon measurements $D_t$, $D_{t-1}$ and ignores all $D_{t-i}|i\geq2$. Hierarchical structuring and conditioning on a larger context have been suggested as ways to add periodicity to Markovian models. These solutions are often computationally intractable or unfeasible [@sadilek_far_2016]. For this reason we do not consider SSMs to be useful for imputing missing data. Nonetheless, they could be of use in filtering noise.

In climate or geological research spatiotemporal imputation methods are often used. For instance, the CUTOFF method estimates missing values using the nearest observed neighbours [ @feng_cutoff:_2014 ]. The authors illustrate their example using rainfall data from gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to impute missing values. The example provided relates to underground water data. Generally these models assume fixed measurement stations (such as rainfall gauging stations).

For this reason they cannot be easily applied to missing mobility tracks. @feng_cutoff:_2014 claim their model could be used to establish mobility patterns. This may be possible by dividing the sample space into rasters. Each raster would be analogous to a measurement station. These artificial stations could "measure" the probability of the individual being there. To our knowledge such models have not been implemented for mobility traces and seems computationally inefficient.

@@

```{r datadetailsTable, echo = F}
datadetails <- readRDS("tempdata/datadescriptives.rds")
datadetails <- datadetails[,-7] #remove last column for now
kable(datadetails,digits = 2,col.names = c("Log duration",
                                           "Logged days",
                                           "Observations",
                                           "Missing days",
                                           "Missing data",
                                           "Mean Accuracy"),
      caption = "Table with descriptives about the data sets used to build the imputation methods. Missing data stands for the proportion of missing 5 minute windows within days that were not missing entirely.")
```

@@

Analyses were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2].

@@
